Online supplement accompanying `The Locally Gaussian Partial Correlation'

### 2021-01-22 lgpc_JBES_supp.Rmd
Paper akseptert, rettet opp i noen små detaljer iht kommentar fra referee

# THE MULTIVARIATE LGPC

@@

\begin{align}
L_{n,b}(\Z_{in}, \Z_{jn}; \rho_b(z_i, z_j)) &= \frac{1}{n}\sum_{t=1}^n K_b(Z_{i,t} - z_i, Z_{j,t} - z_j)\log\left(\psi(Z_{i,t}, Z_{j,t}; \rho_b(z_i, z_j))\right) \nonumber \\  
& \qquad\qquad\qquad - \int K_b(v-z_i, w- z_j)\psi(v,w;\rho_b(z_i, z_j)) \, \di v \di w,
(\#eq:loglikelihood)
\end{align}
where $\Z_{in} = (Z_{i,1}, Z_{i,2}, \ldots, Z_{i,n})$ and $\Z_{jn} = (Z_{j,1}, Z_{j,2}, \ldots, Z_{j,n})$ are the $\Z$-vectors for components $i$ and $j$ respectively, $n$ is the number of observations, $z_i, z_j, v, w \in \mathbb{R}$, $\rho_b = \rho_{ij,b}(z_i,z_j)$ is the local Gaussian correlation parameter, $K_b(v-z_i, w-z_j) = (b_1b_2)^{-1}K(b_1^{-1}(v-z_i))K(b_2^{-1}(w-z_j))$ is a product kernel with bandwidth $b=b_n=(b_1,b_2) = (b_{1n}, b_{2n})$, and $K$ is an non-negative kernel function with compact support such that $\int K(v,w) = 1$. For a fixed bandwidth $b$, $\rho_{0,b}$ is the true local parameter, that is, the minimizer of the local penalty function

\begin{equation}
\int K_b(v-z_i, w-z_j)\big[\psi(v,w; \rho_b(z_i, z_j)) - \log(\psi(v,w; \rho_b(z_i, z_j)))f(v,w)\big] \di v \di w,
(\#eq:penalty)
\end{equation}
where $f$ is the density function of $(Z_i, Z_j)$. Note in particular that $\rho_{0,b} \rightarrow \rho_0$ as $b\rightarrow0$, which is the true local correlation parameter. The reader may consult @tjostheim2013local for more details and regularity conditions in this regard. The corresponding estimate $\widehat \rho = \rho_{n, b}(z_i,z_j)$ is obtained by maximizing the log-likelihood in \@ref(eq:loglikelihood). Note that, writing $\rho$ for $\rho_b$,
\begin{equation}
L_{n,b}(\rho) = \frac{1}{n}\sum_{t=1}^n L^{(b)}(Z_{i,t}, Z_{j,t}; \rho),
\end{equation}
where
\begin{align}
L^{(b)}(Z_{i,t}, Z_{j,t}; \rho) &= K_b(Z_{i,t} - z_i, Z_{j,t} - z_j)\log\left(\psi(Z_{i,t}, Z_{j,t}; \rho_b(z_i, z_j))\right) \nonumber \\  
& \qquad\qquad\qquad - \int K_b(v-z_i, w- z_j)\psi(v,w;\rho_b(z_i, z_j)) \, \di v \di w.
\label{eq:A16}
\end{align}

@@

# ASYMPTOTIC THEORY FOR THE TEST STATISTIC

@@

The results on asymptotic distribution can be proved on several levels. We consider two cases: a) the pairwise simplification where involving the estimates $\rho_{ij,n,b}(z_i,z_j)$ for all possible pairs $(i,j)$ of the $\frho$ vector, and b) the case where $d_1=d_2=d_3=1$, but where $\rho_{ij} = \rho_{ij}(\z)$ is allowed to depend on all three coordinates $\z=(z_1,z_2,z_3)$. We will put the emphasis on a) and only comment briefly on b).

@@

# BOOTSTRAPPING THE TEST STATISTIC

@@

# THE LOCAL BOOTSTRAP

@@

# TWO ADDITIONAL EXAMPLES OF ESTIMATION

@@

# POWER STUDIES

>>>>

This supplement accompanies the article entitled *The Locally Gaussian Partial Correlation*, and contains several technical derivations and discussions. In Section \@ref(multi) we write out the details of the multivariate version of the LGPC. Section \@ref(app-loccor) contains details regarding the proof of Theorem 3.1 in the main article. In Section \@ref(app-asvar) we include some calulations needed to obtain exolicit expressions of the asymptotic variance of the LGPC. Section \@ref(proofpseudo) contains a proof of Theorem 3.2, and in Section \@ref(validity-bootstrap) we present conditions for the validity of the bootstrap when estimating the LGPC. Sections \@ref(asym-test) and \@ref(chap:pitman) present asymptotic theory for the test statistic in the test for conditional independence, while Sections \@ref(boot-test) and \@ref(local-boot) contain various details about bootstrapping the test statistic. In Section \@ref(examples) we include two additional examples -- one simulated and one using real data. Finally, in Section \@ref(power), we present an extension of the power study found in the main article.

# THE MULTIVARIATE LGPC {#multi}

@@

\begin{align}
L_{n,b}(\Z_{in}, \Z_{jn}; \rho_b(z_i, z_j)) &= \frac{1}{n}\sum_{t=1}^n K_b(Z_{i,t} - z_i, Z_{j,t} - z_j)\log\left(\psi(Z_{i,t}, Z_{j,t}; \rho_b(z_i, z_j))\right) \nonumber \\  
& \qquad\qquad\qquad - \int K_b(v-z_i, w- z_j)\psi(v,w;\rho_b(z_i, z_j)) \, \di v \di w,
(\#eq:loglikelihood)
\end{align}
where $\Z_{in} = (Z_{i,1}, Z_{i,2}, \ldots, Z_{i,n})$ and $\Z_{jn} = (Z_{j,1}, Z_{j,2}, \ldots, Z_{j,n})$ are the $\Z$-vectors for components $i$ and $j$ respectively, $n$ is the number of observations, $z_i, z_j, v, w \in \mathbb{R}$, $\rho_b = \rho_{ij,b}(z_i,z_j)$ is the local Gaussian correlation parameter, $K_b(v-z_i, w-z_j) = (b_1b_2)^{-1}K(b_1^{-1}(v-z_i))K(b_2^{-1}(w-z_j))$ is a product kernel with bandwidth $b=b_n=(b_1,b_2) = (b_{1n}, b_{2n})$, and $K$ is an non-negative kernel function with compact support such that $\int K(v,w) = 1$. For a fixed bandwidth $b$, $\rho_{0,b}$ is the true local parameter, that is, the maximizer of the local penalty function

\begin{equation}
\int K_b(v-z_i, w-z_j)\big[\log(\psi(v,w; \rho_b(z_i, z_j)))f(v,w) - \psi(v,w; \rho_b(z_i, z_j))\big] \di v \di w,
(\#eq:penalty)
\end{equation}
where $f$ is the density function of $(Z_i, Z_j)$. Note in particular that $\rho_{0,b} \rightarrow \rho_0$ as $b\rightarrow0$, which is the true local correlation parameter. The reader may consult @tjostheim2013local for more details and regularity conditions in this regard. The corresponding estimate $\widehat \rho = \rho_{n, b}(z_i,z_j)$ is obtained by maximizing the log-likelihood in \@ref(eq:loglikelihood). Note that, writing $\rho$ for $\rho_b$,
\begin{equation}
L_{n,b}(\rho) = \frac{1}{n}\sum_{t=1}^n L^{(b)}(Z_{i,t}, Z_{j,t}; \rho),
\end{equation}
where
\begin{align}
L^{(b)}(Z_{i,t}, Z_{j,t}; \rho) &= K_b(Z_{i,t} - z_i, Z_{j,t} - z_j)\log\left(\psi(Z_{i,t}, Z_{j,t}; \rho_b(z_i, z_j))\right) \nonumber \\  
& \qquad\qquad\qquad - \int K_b(v-z_i, w- z_j)\psi(v,w;\rho_b(z_i, z_j)) \, \di v \di w.
\label{eq:A16}
\end{align}

@@

# ASYMPTOTIC THEORY FOR THE TEST STATISTIC {#asym-test}

@@

The results on asymptotic distribution can be proved on several levels. We consider two cases: a) the pairwise simplification where involving the estimates $\rho_{ij,n,b}(z_i,z_j)$ for all possible pairs $(i,j)$ of the $\frho$ vector, and b) the case where $d_1=d_2=d_3=1$, but where $\rho_{ij} = \rho_{ij}(\z)$ is allowed to depend on all three coordinates $\z=(z_1,z_2,z_3)$. We will put the emphasis on a) and only comment briefly on b). Following Lacal and Tjøstheim (\protect\hyperlink{ref-lacal2018estimating}{2018}) we neglect the asymptotic bias in the asymptotic expressions in Theorems F.2 and F.3. We are then implicitly assuming a suitable bandwidth condition as in Assumption 3 in Theorem 3.1. Deriving explicit expressions for the bias under general circumstances remanins to be worked out.

@@

# BOOTSTRAPPING THE TEST STATISTIC {#boot-test}

@@

# THE LOCAL BOOTSTRAP {#local-boot}

@@

# TWO ADDITIONAL EXAMPLES OF ESTIMATION {#examples}

@@

# POWER STUDIES {#power}

### 2021-01-22 lgpc_JBES_supp.tex
Paper akseptert, rettet opp i noen små detaljer iht kommentar fra referee

\hypertarget{the-multivariate-lgpc}{%
\section{THE MULTIVARIATE LGPC}\label{the-multivariate-lgpc}}

@@

\begin{align}
L_{n,b}(\Z_{in}, \Z_{jn}; \rho_b(z_i, z_j)) &= \frac{1}{n}\sum_{t=1}^n K_b(Z_{i,t} - z_i, Z_{j,t} - z_j)\log\left(\psi(Z_{i,t}, Z_{j,t}; \rho_b(z_i, z_j))\right) \nonumber \\  
& \qquad\qquad\qquad - \int K_b(v-z_i, w- z_j)\psi(v,w;\rho_b(z_i, z_j)) \, \di v \di w,
\label{eq:loglikelihood}
\end{align}
where \(\Z_{in} = (Z_{i,1}, Z_{i,2}, \ldots, Z_{i,n})\) and \(\Z_{jn} = (Z_{j,1}, Z_{j,2}, \ldots, Z_{j,n})\) are the \(\Z\)-vectors for components \(i\) and \(j\) respectively, \(n\) is the number of observations, \(z_i, z_j, v, w \in \mathbb{R}\), \(\rho_b = \rho_{ij,b}(z_i,z_j)\) is the local Gaussian correlation parameter, \(K_b(v-z_i, w-z_j) = (b_1b_2)^{-1}K(b_1^{-1}(v-z_i))K(b_2^{-1}(w-z_j))\) is a product kernel with bandwidth \(b=b_n=(b_1,b_2) = (b_{1n}, b_{2n})\), and \(K\) is an non-negative kernel function with compact support such that \(\int K(v,w) = 1\). For a fixed bandwidth \(b\), \(\rho_{0,b}\) is the true local parameter, that is, the minimizer of the local penalty function

\begin{equation}
\int K_b(v-z_i, w-z_j)\big[\psi(v,w; \rho_b(z_i, z_j)) - \log(\psi(v,w; \rho_b(z_i, z_j)))f(v,w)\big] \di v \di w,
\label{eq:penalty}
\end{equation}
where \(f\) is the density function of \((Z_i, Z_j)\). Note in particular that \(\rho_{0,b} \rightarrow \rho_0\) as \(b\rightarrow0\), which is the true local correlation parameter. The reader may consult Tjøstheim and Hufthammer (\protect\hyperlink{ref-tjostheim2013local}{2013}) for more details and regularity conditions in this regard. The corresponding estimate \(\widehat \rho = \rho_{n, b}(z_i,z_j)\) is obtained by maximizing the log-likelihood in \eqref{eq:loglikelihood}. Note that, writing \(\rho\) for \(\rho_b\),
\begin{equation}
L_{n,b}(\rho) = \frac{1}{n}\sum_{t=1}^n L^{(b)}(Z_{i,t}, Z_{j,t}; \rho),
\end{equation}
where
\begin{align}
L^{(b)}(Z_{i,t}, Z_{j,t}; \rho) &= K_b(Z_{i,t} - z_i, Z_{j,t} - z_j)\log\left(\psi(Z_{i,t}, Z_{j,t}; \rho_b(z_i, z_j))\right) \nonumber \\  
& \qquad\qquad\qquad - \int K_b(v-z_i, w- z_j)\psi(v,w;\rho_b(z_i, z_j)) \, \di v \di w.
\label{eq:A16}
\end{align}

@@

\hypertarget{asymptotic-theory-for-the-test-statistic}{%
\section{ASYMPTOTIC THEORY FOR THE TEST STATISTIC}\label{asymptotic-theory-for-the-test-statistic}}

@@

The results on asymptotic distribution can be proved on several levels. We consider two cases: a) the pairwise simplification where involving the estimates \(\rho_{ij,n,b}(z_i,z_j)\) for all possible pairs \((i,j)\) of the \(\frho\) vector, and b) the case where \(d_1=d_2=d_3=1\), but where \(\rho_{ij} = \rho_{ij}(\z)\) is allowed to depend on all three coordinates \(\z=(z_1,z_2,z_3)\). We will put the emphasis on a) and only comment briefly on b).

@@

\hypertarget{bootstrapping-the-test-statistic}{%
\section{BOOTSTRAPPING THE TEST STATISTIC}\label{bootstrapping-the-test-statistic}}

@@

\hypertarget{the-local-bootstrap}{%
\section{THE LOCAL BOOTSTRAP}\label{the-local-bootstrap}}

@@

\hypertarget{two-additional-examples-of-estimation}{%
\section{TWO ADDITIONAL EXAMPLES OF ESTIMATION}\label{two-additional-examples-of-estimation}}

@@

\hypertarget{power-studies}{%
\section{POWER STUDIES}\label{power-studies}}

>>>>

This supplement accompanies the article entitled \emph{The Locally Gaussian Partial Correlation}, and contains several technical derivations and discussions. In Section \ref{multi} we write out the details of the multivariate version of the LGPC. Section \ref{app-loccor} contains details regarding the proof of Theorem 3.1 in the main article. In Section \ref{app-asvar} we include some calulations needed to obtain exolicit expressions of the asymptotic variance of the LGPC. Section \ref{proofpseudo} contains a proof of Theorem 3.2, and in Section \ref{validity-bootstrap} we present conditions for the validity of the bootstrap when estimating the LGPC. Sections \ref{asym-test} and \ref{chap:pitman} present asymptotic theory for the test statistic in the test for conditional independence, while Sections \ref{boot-test} and \ref{local-boot} contain various details about bootstrapping the test statistic. In Section \ref{examples} we include two additional examples -- one simulated and one using real data. Finally, in Section \ref{power}, we present an extension of the power study found in the main article.

\hypertarget{multi}{%
\section{THE MULTIVARIATE LGPC}\label{multi}}

@@

\begin{align}
L_{n,b}(\Z_{in}, \Z_{jn}; \rho_b(z_i, z_j)) &= \frac{1}{n}\sum_{t=1}^n K_b(Z_{i,t} - z_i, Z_{j,t} - z_j)\log\left(\psi(Z_{i,t}, Z_{j,t}; \rho_b(z_i, z_j))\right) \nonumber \\  
& \qquad\qquad\qquad - \int K_b(v-z_i, w- z_j)\psi(v,w;\rho_b(z_i, z_j)) \, \di v \di w,
\label{eq:loglikelihood}
\end{align}
where \(\Z_{in} = (Z_{i,1}, Z_{i,2}, \ldots, Z_{i,n})\) and \(\Z_{jn} = (Z_{j,1}, Z_{j,2}, \ldots, Z_{j,n})\) are the \(\Z\)-vectors for components \(i\) and \(j\) respectively, \(n\) is the number of observations, \(z_i, z_j, v, w \in \mathbb{R}\), \(\rho_b = \rho_{ij,b}(z_i,z_j)\) is the local Gaussian correlation parameter, \(K_b(v-z_i, w-z_j) = (b_1b_2)^{-1}K(b_1^{-1}(v-z_i))K(b_2^{-1}(w-z_j))\) is a product kernel with bandwidth \(b=b_n=(b_1,b_2) = (b_{1n}, b_{2n})\), and \(K\) is an non-negative kernel function with compact support such that \(\int K(v,w) = 1\). For a fixed bandwidth \(b\), \(\rho_{0,b}\) is the true local parameter, that is, the maximizer of the local penalty function

\begin{equation}
\int K_b(v-z_i, w-z_j)\big[\log(\psi(v,w; \rho_b(z_i, z_j)))f(v,w) - \psi(v,w; \rho_b(z_i, z_j))\big] \di v \di w,
\label{eq:penalty}
\end{equation}
where \(f\) is the density function of \((Z_i, Z_j)\). Note in particular that \(\rho_{0,b} \rightarrow \rho_0\) as \(b\rightarrow0\), which is the true local correlation parameter. The reader may consult Tjøstheim and Hufthammer (\protect\hyperlink{ref-tjostheim2013local}{2013}) for more details and regularity conditions in this regard. The corresponding estimate \(\widehat \rho = \rho_{n, b}(z_i,z_j)\) is obtained by maximizing the log-likelihood in \eqref{eq:loglikelihood}. Note that, writing \(\rho\) for \(\rho_b\),
\begin{equation}
L_{n,b}(\rho) = \frac{1}{n}\sum_{t=1}^n L^{(b)}(Z_{i,t}, Z_{j,t}; \rho),
\end{equation}
where
\begin{align}
L^{(b)}(Z_{i,t}, Z_{j,t}; \rho) &= K_b(Z_{i,t} - z_i, Z_{j,t} - z_j)\log\left(\psi(Z_{i,t}, Z_{j,t}; \rho_b(z_i, z_j))\right) \nonumber \\  
& \qquad\qquad\qquad - \int K_b(v-z_i, w- z_j)\psi(v,w;\rho_b(z_i, z_j)) \, \di v \di w.
\label{eq:A16}
\end{align}

@@

\hypertarget{asym-test}{%
\section{ASYMPTOTIC THEORY FOR THE TEST STATISTIC}\label{asym-test}}

@@

The results on asymptotic distribution can be proved on several levels. We consider two cases: a) the pairwise simplification where involving the estimates \(\rho_{ij,n,b}(z_i,z_j)\) for all possible pairs \((i,j)\) of the \(\frho\) vector, and b) the case where \(d_1=d_2=d_3=1\), but where \(\rho_{ij} = \rho_{ij}(\z)\) is allowed to depend on all three coordinates \(\z=(z_1,z_2,z_3)\). We will put the emphasis on a) and only comment briefly on b). Following Lacal and Tjøstheim (\protect\hyperlink{ref-lacal2018estimating}{2018}) we neglect the asymptotic bias in the asymptotic expressions in Theorems F.2 and F.3. We are then implicitly assuming a suitable bandwidth condition as in Assumption 3 in Theorem 3.1. Deriving explicit expressions for the bias under general circumstances remanins to be worked out.

@@

\hypertarget{boot-test}{%
\section{BOOTSTRAPPING THE TEST STATISTIC}\label{boot-test}}

@@

\hypertarget{local-boot}{%
\section{THE LOCAL BOOTSTRAP}\label{local-boot}}

@@

\hypertarget{examples}{%
\section{TWO ADDITIONAL EXAMPLES OF ESTIMATION}\label{examples}}

@@

\hypertarget{power}{%
\section{POWER STUDIES}\label{power}}
